---
title: "Appendix to the conversion of non-standard h5 files."
output:
  html_document:
    df_print: paged
    highlight: kate
    theme:
      version: 4
      code_font: 
        google: JetBrains Mono
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

# Appendix - inner workings

This section describes in more detail the steps and decisions made when processing a 
non standard H5 file.

## setup

Open the h5 file in read mode, and extract all required slots to vectors. The `[]`
ensures we're getting a vector.

```{r}

h5 <- H5File$new(filename, mode = "r")

counts <- h5[[counts_slot]][]
genes <- h5[[genes_slot]][]
barcodes <- h5[[barcodes_slot]][]

gene_ids <- h5[[gene_ids_slot]][]
gene_names <- h5[[gene_names_slot]][]

```

Barcode indices might be too big, which breaks the creation of the sparse matrix.
So we rank them, to convert to a smaller representation. The `ties.method = "dense"` argument
ensures identical barcodes get the same rank.

```{r}
r_barcodes <- data.table::frankv(barcodes, ties.method = "dense")
```

## matrix creation

Create sparse matrix. Care should be taken with indexing. Depending on the H5 file
indices might be 1 or 0 based. One way is to check for the min of the genes and 
barcodes vectors (which are the indices for the sparse matrix.). If one of them 
contains a zero, then indexing is zero based (the other way around is not necessarily
true, but the worst that can happen is not having an empty row and column.)

```{r}

if(min(genes) == 0 || min(barcodes) == 0) {index1 <- F} else {index1 <- T}

sparse_matrix <-
  sparseMatrix(
    i = genes,
    j = r_barcodes,
    x = counts,
    repr = "C",
    index1 = index1
  )

```

Write the counts to file, we convert the barcodes to a string, because barcodes
are usually nucleotide sequences stored as strings.

```{r}
DropletUtils::write10xCounts(
  path = "./out",
  x = sparse_matrix,
  barcodes = paste0("cell_",  colnames(sparse_matrix)),
  gene.id = gene_ids,
  gene.symbol = gene_names,
  version = "3"
)
```
