---
title: "Convert hdf5 files to features/barcodes/matrix"
output:
  html_document:
    df_print: paged
    highlight: kate
    theme:
      version: 4
      code_font: 
        google: JetBrains Mono
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---


# converting standard h5 to features/barcodes/matrix

Standard 10x hdf5 files can be processed using functionality already implemented
in the `Seurat` package. It should work out of the box for cellranger output. 
If this were to fail, refer to the Non-Standard HDF5 file section of this document.

## libraries

We need to have Seurat and DropletUtils installed. 
Seurat can be installed from CRAN, and [DropletUtils is available on Bioconductor](https://bioconductor.org/packages/release/bioc/html/DropletUtils.html)

```{r}
library(Seurat)
library(DropletUtils)
```

## processing

Set the data_dir to the folder that contains the h5 files. After that, we create
a list of all H5 files in the directory, which will be converted.

```{r}
data_dir <- "./"
setwd(data_dir)
h5_files <- list.files(data_dir, pattern = "*h5$")
```

Create an output directory, to store the converted files.

```{r}
output_dir <- "out"
dir.create(output_dir)
```

Convert the H5 files. The sample_name is going to be the folder name for each sample,
feel free to edit as desired.

```{r}

for (file in h5_files) {
  
  # make sample names, removing .h5
  sample_name <- sub("\\.h5$", "", basename(file))
  sample_path <- file.path(output_dir, sample_name)

  # to show progress
  print(sample_name)
  
  # load the count matrices
  gene_names <- rownames(Seurat::Read10X_h5(file))
  counts <- Seurat::Read10X_h5(file, use.names = F)
  
  # convert
  DropletUtils::write10xCounts(sample_path, counts, version = "3", gene.symbol = gene_names)
}
```

# Processing non-standard H5 files

Non-standard h5's should be treated with care. In particular, the slots with the counts,
genes and barcodes indices could have different names and be located anywhere inside
the H5. Which means manual inspection is required. You can inspect H5 files 
using the `hdf5r` R package, or using the GUI program [HDFView](https://www.hdfgroup.org/downloads/hdfview/).

## libraries

Need to install `hdf5r`, `data.table` and `Matrix` packages. 

```{r}
library(hdf5r)
library(Matrix)
library(DropletUtils)
```

## define parameters

Define slot names, by inspecting the H5 files to be processed, using either `hdf5r`
or HDFView. The slot names are the paths inside the H5 file that point to different
pieces of information required to convert the files, such as the data, or the gene names.

These slot names MUST be changed before processing. They are specific to each non-standard h5 file.

Lengths of the counts, genes and barcodes slots must be the same. These three are
used to build the sparse count matrix.

- `counts_slot` should point to the actual data.
- `genes_slot` should point to an integer vector with row indices
- `barcodes_slot` should point to an integer vector with column indices

```{r}
counts_slot <- "umi_corrected_reads"
genes_slot <- "gene"
barcodes_slot <- "barcode"
```

The gene_names and ids should be the same length and most likely smaller than the
counts/genes/barcodes slots. These are the gene labels, used when creating the 
10x files.

Like the previous slots, these MUST be renamed according to the structure of the specific h5 file being processed.

- `gene_ids_slot` should point to a character vector of gene ids.
- `gene_names_slot` should point to a character vector of gene symbols

```{r}
gene_ids_slot <- "gene_ids"
gene_names_slot <- "gene_names"
```

## Bulk processing

Use this section to bulk process h5 files. If a deeper understanding of the workings
is required, refer to the appendix section at the end of the document.

Set the data_dir to the folder that contains the h5 files. After that, we create
a list of all H5 files in the directory, which will be converted.

```{r}
data_dir <- "./"
setwd(data_dir)
h5_files <- list.files(data_dir, pattern = "*h5$")
```

Create an output directory, to store the converted files.

```{r}
output_dir <- file.path(data_dir, "out")
dir.create(output_dir)
```

### required functions

These are the functions that do the work.

```{r}
extract_slots <- function(h5_path) {
  h5 <- H5File$new(h5_path, mode = "r")
  
  counts <- h5[[counts_slot]][]
  genes <- h5[[genes_slot]][]
  barcodes <- h5[[barcodes_slot]][]
  
  gene_ids <- h5[[gene_ids_slot]][]
  gene_names <- h5[[gene_names_slot]][]
  
  r_barcodes <- data.table::frankv(barcodes, ties.method = "dense")
  
  if(min(genes) == 0 || min(barcodes) == 0) {index1 <- F} else {index1 <- T}
  
  return(
    list(
      "counts" = counts,
      "genes" = genes,
      "barcodes" = barcodes,
      "r_barcodes" = r_barcodes,
      "gene_ids" = gene_ids,
      "gene_names" = gene_names,
      "index1" = index1
    )
  )
}

build_sparse_matrix <- function(slots) {
  sparse_matrix <-
    sparseMatrix(
      i = slots[["genes"]],
      j = slots[["r_barcodes"]],
      x = slots[["counts"]],
      repr = "C",
      index1 = slots[["index1"]]
    )
  
  return(sparse_matrix)
}

```

### processing

Convert all the h5 files detected and stored in the `h5_files` variable.
The sample_name is going to be the folder name for each sample, feel free to edit as desired.

```{r}
for (file in h5_files) {
  print(file)
  
  # make sample names, removing .h5
  sample_name <- sub("\\.h5$", "", basename(file))
  sample_path <- file.path(output_dir, sample_name)

  # to show progress
  print(sample_name)
  
  # read h5 files and build sparse matrix
  slots <- extract_slots(file)
  counts <- build_sparse_matrix(slots)
  
  # write to files.
  DropletUtils::write10xCounts(sample_path,
                               counts,
                               barcodes = paste0("cell_", unique(slots[["barcodes"]])),
                               gene.id = slots[["gene_ids"]],
                               gene.symbol = slots[["gene_names"]],
                               version = "3")
}
```
