---
title: "Removing cellsets from experiment"
output:
  html_document:
    df_print: paged
    highlight: kate
    theme:
      version: 4
      code_font: 
        google: JetBrains Mono
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r}
source("utils.R")
source("cellset_extraction.R")
```

Download data using `biomage experiment download -i production -e EXP ID --all`
Set the `data_dir` var to point to the dir the experiment was downloaded to.

```{r}

data_dir <- "dir/to/downloaded/data"

scdata <- readRDS(file.path(data_dir, "raw_r.rds"))
cellsets <- jsonlite::read_json(file.path(data_dir, "cellsets.json"))
```


Create list of cellset coordinates to extract. Follow `get_cellset_cell_ids` docs
for reference. In this example, we will remove louvain clusters (index 1) 12 and 13,

```{r}
cellsets_to_extract <- list(c(1,13), c(1,14))
```

Extract the cell_ids, and create the vector of cells to remove from the seurat object.
But we need the cell_ids of cells to keep, so we subset the meta.data table, removing the
cells to remove.

```{r}
cell_ids <- get_many_cellset_cell_ids(cellsets, cellsets_to_extract)
cells_to_remove <- match(cell_ids, scdata@meta.data$cells_id)
meta_data_subset <- scdata@meta.data[-cells_to_remove, ]

```

We subset the seurat object, using the pipeline function, removing the cells to remove.
Then we add the sample names, reading the `sample_mapping.json` file, and split
the seurat object using the new `sample_name` column, since we need to recreate the original
samples, minus the removed cells.

```{r}
sub_scdata <- pipeline::subset_ids(scdata, meta_data_subset$cells_id)
sub_scdata <- add_sample_names(sub_scdata, file.path(data_dir, "sample_mapping.json"))
sub_scdata_list <- SplitObject(sub_scdata, split.by = "sample_name")
```

Now we extract the counts and create the features tibble (all feature files will be the same.
Splitting a seurat object only separates cells. All genes are kept for all splits.)

```{r}
sub_scdata_counts <- lapply(sub_scdata_list, extract_counts)
features <-
  tibble(X1 = rownames(sub_scdata_list[[1]]),
         X2 = extract_gene_symbols(sub_scdata_list[[1]]))

tab <- list(counts = sub_scdata_counts,
            sample = names(sub_scdata_counts))
```


We pass the table containing the counts and sample names, the `features` and the `data_dir`
to the `call_writecounts` function, which will output 10x files to a newly created `out/`
subdirectory inside the `data_dir.`

```{r}
tab %>% 
  purrr::pwalk(call_writecounts, features = features,  data_dir = data_dir) 

```

